---
title: "Health and Economic Impact of Extreme Weather Events"
author: "Dave Yachabach"
date: "10/14/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require(data.table)     # fread loads MUCH faster than read.csv
require(R.utils)        # Required by fread()
require(ggplot2)
```

# Synopsis
This document describes the health and economic impact of extreme weather events.  The Health impact is measured in terms of casualties.  Casualties are defined as death or injury as a result of a weather event.  

Economic impact is expressed in Billions of dollars and is the combination of Property Damage and Crop Damage.  

## Data Processing  
The data for this document was originally obtained from the National Oceanic and Atmospheric Administration (NOAA) and was made available to this project through Coursera at [https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).  

```{r Required_Packages}

# Load Data
bz2FilNm <- "repdata_data_StormData.csv.bz2"
dt <- fread(bz2FilNm)

# Get entries that caused casualties or economic impact
dtImpactData <- dt[FATALITIES > 0 | INJURIES > 0 | PROPDMG > 0 | CROPDMG>0,.(FATALITIES, INJURIES,PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP), by = "EVTYPE"]

# correct the value of the damages to Billions
# Correct the "K"
dtImpactData[PROPDMGEXP == "K",]$PROPDMG <- dtImpactData[PROPDMGEXP == "K",]$PROPDMG/1000000
dtImpactData[CROPDMGEXP == "K",]$CROPDMG <- dtImpactData[CROPDMGEXP == "K",]$CROPDMG/1000000

# Correct the "M"
dtImpactData[PROPDMGEXP == "M",]$PROPDMG <- dtImpactData[PROPDMGEXP == "M",]$PROPDMG/1000
dtImpactData[CROPDMGEXP == "M",]$CROPDMG <- dtImpactData[CROPDMGEXP == "M",]$CROPDMG/1000

# Sum to limit the rows
# 
dtImpactTotals <- dtImpactData[,.(CASUALTIES = sum(FATALITIES,INJURIES), ECONOMIC = sum(PROPDMG, CROPDMG)), by = "EVTYPE"]

```
### Manipulations  
Weather event data is filtered to only those events with economic or public health impact.  

- Public health impact = Casualties (Fatality + Injury)  

- Economic impact = Economic (Property + Crop damage)  

This results in `r length(dtImpactTotals[,EVTYPE])` categories of weather events.  

### Cleaning and Consolidating  
Economic data were expressed in two fields:  

- PROPDMG and  CROPDMG: Numeric - Indicates Amount  

- PROPDMGEXP and CROPDMGEXP: Character - "K", "M", or "B"  

Each entry was checked and corrected to "Billions".  

Data includes multiple names for similar storm types.  These names were consolidated into 11 event categories. The table in Appendix 1 shows how public health threats were consolidated.

Additionally, Fatalities and Injuries were consolidated into Casualties; Property and Crop damage were consolidated into Economic.  

```{r EventConsolidation}
# Build lists of event descriptions
# 

lstTSTMWEvent <- grep("THUND|TST", dtImpactTotals [,EVTYPE], value = TRUE)
lstTRNDOEvent <- grep("TORN|SPOUT|FUNN", dtImpactTotals $EVTYPE, value = TRUE)
lstWINDEvent <- grep("^HIGH WIND$|^HIGH WINDS*$", dtImpactTotals $EVTYPE, value = TRUE)
lstLTNGEvent <- grep("^[L].*G.*G", dtImpactTotals [,EVTYPE], value = TRUE)
lstHEATEvent <- grep("HEAT|WARM", dtImpactTotals [,EVTYPE], value = TRUE)
lstHRCNEvent <- grep("HUR|Hur|Typ|TYP|TROP|Trop", dtImpactTotals $EVTYPE, value = TRUE)
lstFLOODEvent <- grep("FLOO|Floo|FLA|Fla", dtImpactTotals $EVTYPE, value = TRUE)
lstCOLDEvent <- grep("COLD.*|Cold", dtImpactTotals $EVTYPE, value = TRUE)
lstWINTEREvent <- grep("WINT|Wint", dtImpactTotals $EVTYPE, value = TRUE)
lstSNOWEvent <- grep("SNOW", dtImpactTotals $EVTYPE, value = TRUE)
lstHAILEvent <- grep("HAIL", dtImpactTotals $EVTYPE, value = TRUE)
lstFIREEvent <- grep("FIRE", dtImpactTotals $EVTYPE, value = TRUE)
lstTIDEEvent <- grep("CURR|TIDE", dtImpactTotals $EVTYPE, value = TRUE)
lstFOGEvent <- grep("FOG", dtImpactTotals $EVTYPE, value = TRUE)
lstBLZEvent <- grep("BLI", dtImpactTotals $EVTYPE, value = TRUE)
lstICEEvent <- grep("ICE", dtImpactTotals $EVTYPE, value = TRUE)
lstSEAEvent <- grep("SEAS$|TIDE|CURR", dtImpactTotals $EVTYPE, value = TRUE)

# Then apply those descriptions to EVTYPE to summarize into EVENT categories

dtImpactTotals [,EVENT := "OTHER"]
dtImpactTotals [EVTYPE %in% lstTRNDOEvent, ]$EVENT = "TORNADO"
dtImpactTotals [EVTYPE %in% lstTSTMWEvent, ]$EVENT = "TSTMS/HAIL/LIGHTNING"
dtImpactTotals [EVTYPE %in% lstHAILEvent, ]$EVENT = "TSTMS/HAIL/LIGHTNING"
dtImpactTotals [EVTYPE %in% lstWINDEvent,]$EVENT = "WINDS"
dtImpactTotals [EVTYPE %in% lstLTNGEvent, ]$EVENT = "TSTMS/HAIL/LIGHTNING"
dtImpactTotals [EVTYPE %in% lstHEATEvent, ]$EVENT = "HEAT"
dtImpactTotals [EVTYPE %in% lstHRCNEvent, ]$EVENT = "HURRICANE"
dtImpactTotals [EVTYPE %in% lstFLOODEvent, ]$EVENT = "FLOOD"
dtImpactTotals [EVTYPE %in% lstCOLDEvent, ]$EVENT = "WINTER/COLD/SNOW"
dtImpactTotals [EVTYPE %in% lstWINTEREvent, ]$EVENT = "WINTER/COLD/SNOW"
dtImpactTotals [EVTYPE %in% lstSNOWEvent, ]$EVENT = "WINTER/COLD/SNOW"
dtImpactTotals [EVTYPE %in% lstBLZEvent,]$EVENT = "WINTER/COLD/SNOW"
dtImpactTotals [EVTYPE %in% lstICEEvent,]$EVENT = "WINTER/COLD/SNOW"
dtImpactTotals [EVTYPE %in% lstFIREEvent,]$EVENT = "WILDFIRE"
dtImpactTotals [EVTYPE %in% lstTIDEEvent,]$EVENT = "SEAS/TIDES"
dtImpactTotals [EVTYPE %in% lstFOGEvent,]$EVENT = "FOG"

```
# Results  
## Public Health Impact
The graph below shows the relative public health hazard of the top 5 consolidated weather event categories since 1950:  

```{r HealthTop5Graph}

# We need top 5 casualty events
dtHealthTop5 <- dtImpactTotals[,.(CASUALTIES = sum(CASUALTIES)), by = "EVENT"]
setorder(dtHealthTop5, -CASUALTIES)
dtHealthTop5 <- dtHealthTop5[1:5,]
dtHealthTop5

# Now plot
g <- ggplot(data = dtHealthTop5)
ga <- g + aes(x = dtHealthTop5$EVENT, y = dtHealthTop5$CASUALTIES)
gag <- ga + geom_bar(stat = "identity", fill = "blue")
gagl <- gag + ggtitle ("Top 5 Most Dangerous Weather Events") + xlab("Event") + ylab("Casualties") + theme(legend.position = "none", axis.text.x = element_text(angle = 90))
gagl + scale_x_discrete(limits = dtHealthTop5$EVENT)

```

## Economic Impact  
The graph below shows the relative economic hazard of the top 5 consolidated weather event categories since 1950:

```{r EconTop5Graph}
# Get top 5 Economic events
dtEconTop5 <- dtImpactTotals[,.(ECONOMIC_Blns = sum(ECONOMIC)), by = "EVENT"]
setorder(dtEconTop5, -ECONOMIC_Blns)
dtEconTop5 <- dtEconTop5[1:5,]
dtEconTop5

# Now plot
g <- ggplot(data = dtEconTop5)
ga <- g + aes(x = dtEconTop5$EVENT, y = dtEconTop5$ECONOMIC_Blns)
gag <- ga + geom_bar(stat = "identity", fill = "green")
gagl <- gag + ggtitle ("Top 5 Most Costly Weather Events") + xlab("Event") + ylab("Economic (Billions $)") + theme(legend.position = "none", axis.text.x = element_text(angle = 90))
gagl + scale_x_discrete(limits = dtEconTop5$EVENT)

```


# Appendix 1  
EVENT|EVTYPE|  
-|-|  
TORNADO|`r dtImpactTotals [EVENT == "TORNADO",EVTYPE]`|  
TSTMS/HAIL/LIGHTNING|`r dtImpactTotals [EVENT == "TSTMS/HAIL/LIGHTNING",EVTYPE]`|  
HEAT|`r dtImpactTotals [EVENT == "HEAT",EVTYPE]`|  
FLOOD|`r dtImpactTotals [EVENT == "FLOOD",EVTYPE]`|  
WINTER/COLD/SNOW|`r dtImpactTotals [EVENT == "WINTER/COLD/SNOW",EVTYPE]`|  
HURRICANE|`r dtImpactTotals [EVENT == "HURRICANE",EVTYPE]`|  
WINDS|`r dtImpactTotals [EVENT == "WINDS",EVTYPE]`| 
WILDFIRE|`r dtImpactTotals [EVENT == "WILDFIRE",EVTYPE]`| 
SEAS/TIDES|`r dtImpactTotals [EVENT == "SEAS/TIDES",EVTYPE]`| 
FOG|`r dtImpactTotals [EVENT == "FOG",EVTYPE]`| 
OTHER|`r dtImpactTotals [EVENT == "OTHER",EVTYPE]`| 


